\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-22 analysis/manuscript.tex
first draft discussion

With the introduction of central funding mechanisms for publication fees
in some European countries since 2012, an increase in hybrid open access
could be observed (Björk, 2017; Huang et al., 2020; Jubb et al., 2017;
Piwowar et al., 2018). For example, studying university output,
Robinson-Garcia et al. (2020) estimated a median uptake of 7.1\% in the
period 2014-2017. In particular, British (17\%), Austrian (15\%) and
Dutch (13\%) universities contributed to this trend. However, this shift
in funding policy towards hybrid open access also added to the overall
cost of publishing, which includes subscription spending and the
administrative efforts required to handle payments (Pinfield et al.,
2016). Moreover, established large commercial publishers, which already
dominated the publishing market (Larivière et al., 2015),
disproportionately benefited from hybrid open access funding in
comparison to full open access publishers (Butler et al., 2023; Jahn \&
Tullney, 2016; Shu \& Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing, as well as to increase publisher-provided
immediate open access (Björk \& Solomon, 2014; Schimmer et al., 2015).
These considerations resulted into transformative agreements\footnote{In
  this paper I use the term ``transformative agreement'', addressing
  also offsetting, read-and-publish or publish-and-read deals, and other
  variants (Borrego et al., 2020; Hinchliffe, 2019). Although the term
  is critised as misleading and not useful to describe these agreements
  between library consortia and publishers (Babini et al., 2022), it is
  widely used in policy discussions and in the research literature.},
which cover a broad range of contracts between library consortia and
publishers from mid-2010s onwards where institutional spending for
subscriptions and open access publishing are considered together
(Borrego et al., 2020; Hinchliffe, 2019). Transformative agreements seek
to control costs while allowing a transitional phase for publishing more
open access articles. Similar to big deals, transformative agreements
mainly bundle hybrid and subscription-only journals from commercial
publishers, but aim at a higher degree of transparency than previous big
deals, where contracts including payments were confidential (Bergstrom
et al., 2014).

@@

By the end of 2023, many transformative agreements were implemented, but
interim outcomes are mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both full open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilization of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuate market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality, because transformative agreements focus on pay to publish
open access mainly targets institutions from high-income countries,
furthering a questionable journal prestige culture (Babini et al.,
2022). Besides, an editorial-board resignation raised concerns that
publishers' desire to maximize journal publication volume ``without
regard to quality'' is a consequence of transformative agreements
(Rasmussen, 2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended to only support
agreements for publishing in full open access journals \footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements by the end of 2024 (Liverpool, 2023). The
consortium also removed most hybrid journals from its Transformative
Journal program in 2023 due to publishers' failure to meet self-defined
open access growth targets (Brainard, 2023). In contrast, the German
DEAL consortium signed a five-year transformative agreement with
Elsevier starting in 2024, while also renewing its contracts with
Springer Nature and Wiley until the end of 2028. At the same time, the
Colombia Consortium signed the first transformative agreements in Latin
America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this uptake can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while large-scale studies
relied on self-reported agreement data (Moskovkin et al., 2022), or used
APC pricing lists (Shu \& Larivière, 2023). Particularly, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements.

The present study aims to address these limitations by combining
multiple openly available data sources to determine the open access
uptake in hybrid journals, while distinguishing between open access
through transformative agreements and other means. With this novel and
open approach, this the first large-scale analysis aims to answer the
following questions:

@@

This study combines data from multiple publicly available data sources
as diagrammatically shown in Figure \ref{fig:data_workflow}. Initially,
transformative agreement data retrieved from the cOAlition S Journal
Checker Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identification of hybrid journals by excluding full
open access journals, Crossref served as the primary data source for
article-level metadata including Creative Commons (CC) license
information to indicate open access availability on publisher websites.
To determine open access articles published through transformative
agreements, first author affiliations from OpenAlex (Priem et al., 2022)
were subsequently linked to eligible institutions according to the
transformative agreement data. In the following, these steps are
described in more detail.

@@

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017). Comparing Crossref license coverage with OpenAlex,
which re-uses open access evidence from Unpaywall, a widely used open
access discovery service that also parses journal websites for open
content licenses (Piwowar et al., 2018), highlighted ongoing challenges
to identify hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
with CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The biggest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers only differ slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable difference
could be furthermore observed among some publishers that presumably did
not provide CC license information to Crossref including AIP Publishing,
American Physiological Society, Emerald and the Royal Society. Crossref
license metadata was more complete with regard to articles from the
publisher Wiley and American Chemical Society. Finally, inconsistent
open access status information in previous OpenAlex versions was
observed (Jahn et al., 2023). After reporting it to OpenAlex, according
to the release notes, fixing this issue was still ongoing, which might
also explain part of the discrepancy.

@@

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

@@

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable growth in open access
provision through other means, most likely publication fees being not
invoiced through transformative agreements, which increased from 4.1\%
(n = 62,625) in 2018 to 6.3\% (n = 105,896).

@@

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year with a focus on open
articles enabled by transformative agreements. Although all publishers
show a general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that, in particular, Wiley
experienced a substantial increase in its open access share from 5.9\%
(n = 11,628) in 2018 to 26\% (n = 53,503) in 2022, representing an
4.5-fold increase. In contrast, Elsevier's hybrid journals demonstrated
a more modest increase, from 3.3\% (n = 16,872) in 2018 to 10\% (n =
60,821) in 2022, which is a relatively low open access share compared to
the general trend. In 2018, Springer Nature had the largest open access
proportion among the three publishers of 8.4\% (n = 19,701), but
experienced a relatively slower growth, resulting in 18\% (n = 52,616)
of articles being open access in Springer Nature hybrid journals in
2022.

@@

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa as of 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded an low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals. While OECD publication output
in Elsevier's Physical Sciences journals declined from 112,822 articles
in 2018 to 103,766 in 2022, BRICS output increased from 104,654 to
171,713 in the same five-year period. Furthermore, OECD publication
output in Health Science Journals and Life Science journals stagnated
across the investigated hybrid journal portfolios after a peak in 2020.

@@

To illustrate the situation in 2022, Figure \ref{fig:country_patch}B
compares total publication output with the number of open access
articles. With 391,530 articles, China was the most productive country,
followed by the United States (268,965 articles) and India (87,428
articles). In contrast, West and Nord European countries published a
considerable high number of open access articles, mainly due to
transformative agreements. Particularly, Germany, Great Britain, the
Netherlands, Sweden, Switzerland and Spain recorded an above-average
open access share as indicated by the linear trend line. As represented
by the point size, as well as it can been seen in Figure
\ref{fig:country_patch}C, transformative agreements contributed to these
market positions. Interestingly, the United States had a notable open
access market share of 15\%, although transformative agreements
contributed to a lesser extent. Similarly, China's open access market
share of 7.2\% in 2022 was comparable to that of the Netherlands, which
was 7.1\%.

Figure \ref{fig:country_top_20_plot} illustrates the development of
hybrid open access from 2018 to 2022, highlighting the top 20 most
productive countries in terms of articles published in hybrid journals
that were included in transformative agreements over the five-year
period. Notably, The Netherlands (27\%), Sweden (24\%), Poland (17\%)
and Great Britain (17\%)) exhibited a relatively high level of uptake in
2018 which continued to increase in the following years. In 2022, Sweden
had the highest proportion of open access relative to its publication
output (78\%), followed by the Netherlands (67\%) and Switzerland
(57\%), with these countries benefiting from transformative agreements.
In Germany, however, hybrid open access only began to increase from 2019
onwards after the successful negotiation of nationwide agreements with
Wiley (July 2019) and Springer Nature (January 2020). Prior to this,
only a few organisations had agreements in place, for the example the
Max Planck Society with Springer Compact.

@@

Since 2021, there has been a general trend towards hybrid open access
among many high-income countries, primarily driven by transformative
agreements. However, proliferation of transformative agreements differed
across these countries. For instance, Germany successfully negotiated an
agreement with Elsevier not until 2023. Additionally, publication limits
or eligibility criteria for institutions and article types may explain
why even countries with widespread agreement implementation do not
achieve 100\% hybrid open access. Interestingly, in Japan and the US
other options than transformative agreements were the main driver for
the increase in hybrid open access. Once again, the graph highlights
countries with low hybrid open access, particularly non-OECD countries,
where only a few or no agreements were in place.

@@

The primary aim of this study was to investigate the adoption of open
access in hybrid journals in transformative agreements, which were
introduced as a temporal means to support the transition of
subscription-based academic publishing to full open access. This study
presents a novel approach based on open data, which leverages metadata
on over 700 agreements and nine million journal articles to estimate the
extent to which transformation agreement contribute to the transition of
this journal business model based on institutional affiliation. The
results highlight a strong growth in open access between 2018 and 2022,
particularly driven by an increasing number of institutions that had
transformative agreements in place. Yet, the majority of research
literature published in hybrid journals in this five-year period
remained behind a publisher's paywall. Growth in the adoption of open
access in hybrid journals, in particular through transformative
agreements, can be largely attributed to three large commercial
publishers -- Elsevier, Springer Nature, and Wiley -- but varies
substantially across journals, publishers, subjects, and country
affiliations. Despite the limitations of the data used, the findings
indicate that the current level of implementation of transformative
agreements is insufficient to bring about a large-scale transition to
fully open access.

A key finding of the analysis is that transformative agreements increase
market concentration. Specifically, the three largest commercial
publishers---Elsevier, Springer Nature, and Wiley---have increased their
market share in the context of open access, particularly in regard to
open access provided through transformative agreements. Together, the
three publishers accounted for three-fourths of open access articles
through transformative agreements, despite publishing less than half of
the total publication volume published between 2018 and 2022. This
observation aligns with previous research on shifts in the publishing
market following the introduction of funding opportunities for hybrid
open access (Butler et al., 2023; Jahn \& Tullney, 2016; Shu \&
Larivière, 2023). Additionally, the results confirm previously observed
variations by publisher, with Elsevier exhibiting a different
development than Springer Nature and Wiley (Butler et al., 2023; Jahn et
al., 2022). However, it must be noted that the focus of transformative
agreements on publishers with large hybrid journal portfolios is
intentional (Campbell et al., 2022). Because of transformative
agreements with a few large publishers, national consortia were able to
substantially increased their country's annual open access article
volume (Huang et al., 2020; Pinhasi et al., 2021; Taubert et al., 2023;
Wenaas, 2022).

Moreover, this study presents varying levels of open access uptake
through transformative agreements across journals, which can be
attributed to the alignment of authors' affiliations and the
availability of such agreements at their institutions. In line with
previous research findings (Butler et al., 2023; Jahn et al., 2022),
high uptake rates were observed across hybrid journals in the Social
Sciences and Humanities. However, it is important to emphasise that
these hybrid journals do not encompass the entire field. For example, a
study by Khanna et al. (2022) found that approximately 60\% of journals
utilizing the open-source publishing platform Open Journal Systems (OJS)
fall within the Social Sciences and Humanities. In these fields, the
majority of full open access journals are so-called ``Diamond OA
journals'' that do not charge publication fees (Bosman et al., 2021).
Rather, this result can be more accurately attributed to the substantial
proportion of authors from high-income countries who publish in these
hybrid journals included in transformative agreements, particularly when
journals are of local relevance, such as through the belonging to
national societies or regional research scope.

Surprisingly, the total publication output of high-income countries
belonging to the OECD declined susbtantially between 2018 and 2022,
while the BRICS countries recorded an steady increase. Because the BRICS
expansion can be exclusivley attributed to closed access articles, this
developement has the potential to hinder the transition of academic
journal publishing to full open access through transformative
agreements. In China, the most productive country according to the total
article volume in hybrid journals, (shu und zhang und das indische
paper)

other countries less represented and brazil diamond russia The Eastern
Europe Expanded Collection

but not coherent acorss all countries, in particular in the us.
neverthelles china and us have susbtantial market shares when it comes
to hybrid oa.

There are at least three potential limitations concerning the results of
this study. A first limitation concerns \ldots{} A second potential
limitation is that \ldots{}

These data have some potential intervention implications. forst and
formesot, the supprot recent pocliy decision highlighting a grew in oa
at country-level, but questioning the. if thouse contracts, a
transiliteration is ttoo short and might require a longer
transiltierional period. the growtrh in brics however would require too
impleis unboundling, other payment means and focus on alternative, in
particular innovations like preprinting, diamond oa

In terms of future research, it would be useful to extend the current
findings by examining other oa types, combine oa with mobility.
fianncials tudies are also required to assess claims of cost-sensitive,
here, publisher and publsiher are required to increase transparency. at
the same time, studies needed to assess these claim economically. data
can be improved to entail more, in apritcular consortia could start to
disclose funded articles as , with openapc already a cwors-orucing
platform available. also libraries can contrbute to an improved
catalogiuing of transformative agreements including participating
insitutions, which is aligned with standards used in schoalrly
communications like ror id.

In summary, after more than tow decades, hybrid open access has not
demonstrated a succesful menas. stduying transofrmative agreemnts,
effects can be found at the -level of participating countries, but
overlla uptake remains low, in particular with regard to publications
from brocs countries, which country share rise, exclusively on the
expense of closed access. means

>>>>

With the introduction of central funding mechanisms for publication fees
in some European countries since 2012, an increase in hybrid open access
could be observed (Björk, 2017; Huang et al., 2020; Jubb et al., 2017;
Piwowar et al., 2018). For example, studying university output,
Robinson-Garcia et al. (2020) estimated a median uptake of 7.1\% in the
period 2014-2017. In particular, British (17\%), Austrian (15\%) and
Dutch (13\%) universities stand out. However, this shift in funding
policy towards hybrid open access also added to the overall cost of
publishing, which includes subscription spending and the administrative
efforts required to handle payments (Pinfield et al., 2016). Moreover,
established large commercial publishers, which already dominated the
publishing market (Larivière et al., 2015), disproportionately benefited
from hybrid open access funding in comparison to full open access
publishers (Butler et al., 2023; Jahn \& Tullney, 2016; Shu \&
Larivière, 2023).

As a consequence, libraries and their consortia began to develop
licensing strategies aimed at avoiding such `double dipping' scenarios,
in which well-established commercial publishers gain twice from reading
and open access publishing (Mittermaier, 2015), as well as to increase
publisher-provided immediate open access (Björk \& Solomon, 2014;
Schimmer et al., 2015). These considerations resulted into
transformative agreements\footnote{In this paper I use the term
  ``transformative agreement'', addressing also offsetting,
  read-and-publish or publish-and-read deals, and other variants
  (Borrego et al., 2020; Hinchliffe, 2019). Although the term is
  critised as misleading and not useful to describe the different types
  of open access agreements between library consortia and commercial
  publishers (Babini et al., 2022), it is widely used in policy
  discussions and in the research literature.}, which cover a broad
range of contracts between library consortia and publishers from
mid-2010s onwards where institutional spending for subscriptions and
open access publishing are considered together (Borrego et al., 2020;
Hinchliffe, 2019). Transformative agreements seek to control costs while
allowing a transitional phase for publishing more open access articles.
Similar to big deals, transformative agreements mainly bundle hybrid and
subscription-only journals from commercial publishers, but aim at a
higher degree of transparency than previous big deals, where contracts
including payments were confidential (Bergstrom et al., 2014).

@@

By the end of 2023, many transformative agreements were implemented, but
interim outcomes are mixed. The ESAC Transformative Agreement
Registry\footnote{\url{https://esac-initiative.org/about/transformative-agreements/agreement-registry/}},
the largest source of disclosure, recorded more than 800 transformative
agreements, resulting in up to 900.000 open access articles published in
both full open access and hybrid journals according to the accompanying
ESAC Market Watch\footnote{\url{https://esac-initiative.org/market-watch/}}.
Library consortia reported increased open access volume, streamlined
payment and monitoring procedures, as well as extensive utilization of
open access options by the researchers they serve (Marques \& Stone,
2020; Parmhed \& Säll, 2023; Pinhasi et al., 2020). The ongoing
standardisation of transformative agreements contributed to improved
transparency in terms of contracts and publisher-provided article
metadata (Marques et al., 2019; Pinhasi et al., 2021). However, with the
growing trend toward transformative agreements, continued reliance on
big deals is perceived as problematic, because it perpetuate market
concentration (Butler et al., 2023; Shu \& Larivière, 2023). Whether
transformative agreements lead to reduced pricing remains uncertain
(Borrego, 2023) and a substantial transition of hybrid journals towards
full open access could not be observed (Matthias et al., 2019; Momeni et
al., 2021). The focus on large commercial publishers might also increase
inequality (Klebel \& Ross-Hellauer, 2023), because transformative
agreements focus on pay to publish open access mainly targets
institutions from high-income countries, furthering a questionable
journal prestige culture (Babini et al., 2022). Besides, an
editorial-board resignation raised concerns that publishers' desire to
maximize journal publication volume ``without regard to quality'' is a
consequence of transformative agreements (Rasmussen, 2023).

The controversies surrounding hybrid open access and transformative
agreements have led to varying policy conclusions. For instance, the
Association of Swedish Higher Education Institutions (Sveriges
universitets- och högskoleförbund, SUHF) recommended to only support
agreements for publishing in full open access journals \footnote{\url{https://www.su.se/english/news/open-access-need-to-move-away-from-transformative-agreements-1.683787}}.
Likewise, most cOAlition S funders will discontinue financial support
for transformative agreements by the end of 2024 (Liverpool, 2023). The
consortium also removed most hybrid journals from its Transformative
Journal program in 2023 due to publishers' failure to meet self-defined
open access growth targets (Brainard, 2023). In contrast, the German
DEAL consortium announced a five-year transformative agreement with
Elsevier starting in 2024, while also renewing its contracts with
Springer Nature and Wiley until the end of 2028. At the same time, the
Colombia Consortium signed the first transformative agreements in Latin
America (Muñoz-Vélez et al., 2024).

Despite these controversies around transformative agreements as a means
of transitioning journal publishing to full open access, there is
limited evidence available on the adoption of open access in hybrid
journals, and the extent to which this uptake can be attributed to
transformative agreements. Previous studies have focused on specific
countries (Haucap et al., 2021; Huang et al., 2020; Pölönen et al.,
2020; Taubert et al., 2023; Wenaas, 2022) or publisher portfolios
(Bakker et al., 2024; Fraser et al., 2023; Jahn et al., 2022; Momeni et
al., 2023; Pieper \& Broschinski, 2018), while large-scale studies
relied on self-reported agreement data (Moskovkin et al., 2022), or used
APC pricing lists (Shu \& Larivière, 2023). Particularly, data
availability is a limiting factor when studying the impact of
transformative agreements (Bakker et al., 2024), because bibliometric
databases, even though many allow the retrieval of open access articles
in hybrid journals, do not directly attribute them to specific
transformative agreements. Likewise, article-level open access invoicing
data is only partly available (Jahn et al., 2022).

The present study aims to address these limitations by combining
multiple openly available data sources to determine the open access
uptake in hybrid journals, while distinguishing between open access
through transformative agreements and other means. With this novel and
open approach, this first large-scale analysis aims to answer the
following questions:

@@

This study combines data from multiple publicly available data sources
as shown in Figure \ref{fig:data_workflow}. Initially, transformative
agreement data retrieved from the cOAlition S Journal Checker
Tool\footnote{\url{https://www.coalition-s.org/blog/enabling-accurate-results-within-the-journal-checker-tool/}}
provided information about journal portfolios and participating
institutions. After identification of hybrid journals by excluding full
open access journals, Crossref served as the primary data source for
article-level metadata including Creative Commons (CC) license
information to indicate open access availability on publisher websites.
To determine open access articles published through transformative
agreements, first author affiliations from OpenAlex (Priem et al., 2022)
were subsequently linked to eligible institutions according to the
transformative agreement data. In the following, these steps are
described in more detail.

@@

Open access articles in hybrid journals were identified through Creative
Commons (CC) license information in Crossref metadata. License
information relative to the ``accepted manuscript (AM)'' version were
not considered. Crossref was used for open access identification because
transformative agreements workflows generally require publishers to
deliver CC license information to this DOI registration agency (Geschuhn
\& Stone, 2017).

Comparing Crossref license coverage with OpenAlex, which re-uses open
access evidence from Unpaywall, a widely used open access discovery
service that also parses journal websites for open content licenses
(Piwowar et al., 2018), highlighted ongoing challenges to identify
hybrid open access (Butler et al., 2023; Jahn et al., 2022;
Martín-Martín et al., 2018; Zhang et al., 2022). Here, 742,369 articles
with CC license were retrieved using Crossref, while 950,260 articles
were tagged as ``hybrid'' according to the OpenAlex November 2023
release, which was used throughout this study. The biggest differences
concerned articles published between 2018 and 2020. In 2022, however,
Crossref and OpenAlex open access numbers only differ slightly (249,511
records using Crossref vs.~255,344 in OpenAlex). Notable difference
could be furthermore observed among some publishers that presumably did
not provide CC license information to Crossref including AIP Publishing,
American Physiological Society, Emerald and the Royal Society. Crossref
license metadata was more complete with regard to articles from the
publisher Wiley and American Chemical Society. Finally, inconsistent
open access status information in previous OpenAlex versions was
observed (Jahn et al., 2023). After reporting it to OpenAlex, fixing
this issue was still ongoing according to the release notes, which might
also explain part of the discrepancy.

@@

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

@@

Figure \ref{fig:results_overview}B highlights that the majority of
hybrid open access was made available through transformative agreements
in 2021 and 2022, contributing 58\% of the total open access article
volume in 2022. However, there was also a notable growth in open access
provision through other means, presumably publication fees being not
invoiced through transformative agreements, which increased from 4.1\%
(n = 62,625) in 2018 to 6.3\% (n = 105,896).

@@

Figure \ref{fig:publisher_figure} takes a closer look into the growth of
hybrid open access across publishers by year with a focus on open access
enabled by transformative agreements. Although all publishers show a
general long-term trend towards transformative agreements, Figure
\ref{fig:publisher_figure}A and B indicate that, in particular, Wiley
experienced a substantial increase in its open access share from 5.9\%
(n = 11,628) in 2018 to 26\% (n = 53,503) in 2022, representing an
4.5-fold increase. In contrast, Elsevier's hybrid journals demonstrated
a more modest increase, from 3.3\% (n = 16,872) in 2018 to 10\% (n =
60,821) in 2022, which is a relatively low open access share compared to
the general trend. In 2018, Springer Nature had the largest open access
proportion among the three publishers of 8.4\% (n = 19,701), but
experienced a relatively slower growth, resulting in 18\% (n = 52,616)
of articles being open access in Springer Nature hybrid journals in
2022.

@@

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa as of 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded a low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite the rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals. While OECD publication output
in Elsevier's Physical Sciences journals declined from 112,822 articles
in 2018 to 103,766 in 2022, BRICS output increased from 104,654 to
171,713 in the same five-year period. Furthermore, OECD publication
output in Health Science Journals and Life Science journals stagnated
across the investigated hybrid journal portfolios after a peak in 2020.

@@

To illustrate the situation in 2022, Figure \ref{fig:country_patch}B
compares total publication output with the number of open access
articles. With 391,530 articles, China was the most productive country,
followed by the United States (268,965 articles) and India (87,428
articles). In contrast, West and Nord European countries published a
considerable high number of open access articles. Particularly, Germany,
Great Britain, the Netherlands, Sweden, Switzerland and Spain recorded
an above-average open access share as indicated by the linear trend
line. As represented by the point size, as well as it can been seen in
Figure \ref{fig:country_patch}C, transformative agreements contributed
to these market positions. Interestingly, the United States had a
notable open access market share of 15\%, although transformative
agreements contributed to a lesser extent. Similarly, China's open
access market share of 7.2\% in 2022 was comparable to that of the
Netherlands, which was 7.1\%.

Figure \ref{fig:country_top_20_plot} illustrates the development of
hybrid open access from 2018 to 2022, highlighting the top 20 most
productive countries in terms of articles published in hybrid journals
that were included in transformative agreements over the five-year
period. Notably, The Netherlands (27\%), Sweden (24\%), Poland (17\%)
and Great Britain (17\%) exhibited a relatively high level of uptake in
2018 which continued to increase in the following years. In 2022, Sweden
had the highest proportion of open access relative to its publication
output (78\%), followed by the Netherlands (67\%) and Switzerland
(57\%), with these countries benefiting from transformative agreements.
In Germany, however, hybrid open access only began to increase from 2019
onwards after the successful negotiation of nationwide agreements with
Wiley (July 2019) and Springer Nature (January 2020). Prior to this,
only a few organisations had agreements in place, for the example the
Max Planck Society with Springer Compact.

@@

Since 2021, there has been a general trend towards hybrid open access
among many high-income countries, primarily driven by transformative
agreements. However, proliferation of transformative agreements differed
across these countries. For instance, Germany successfully negotiated an
agreement with Elsevier not until 2023. Additionally, publication limits
or eligibility criteria for institutions and article types may explain
why even countries with widespread agreement implementation like Sweden
or the Nethterlands did not achieve 100\% hybrid open access.
Interestingly, in Japan and the US other options than transformative
agreements were the main driver for the increase in hybrid open access.
Once again, the graph highlights countries with low hybrid open access,
particularly non-OECD countries, where only a few or no agreements were
in place.

@@

The primary aim of this study was to investigate the adoption of open
access in hybrid journals in transformative agreements, which were
introduced as a temporal means to support the transition of
subscription-based academic publishing to full open access. This study
presents a novel approach based on open data, which leverages metadata
on over 700 agreements and nine million journal articles to estimate the
extent to which transformation agreement contribute to the transition of
this journal business model. The results highlight a strong growth in
open access between 2018 and 2022, particularly driven by an increasing
number of institutions that had transformative agreements in place. Yet,
the majority of research literature published in hybrid journals in this
five-year period remained behind a publisher's paywall. Growth in the
adoption of open access in hybrid journals, in particular through
transformative agreements, can be largely attributed to three large
commercial publishers -- Elsevier, Springer Nature, and Wiley -- but
varies substantially across journals, publishers, subjects, and country
affiliations. Despite the limitations of the data used, the findings
indicate that the current level of implementation of transformative
agreements is insufficient to bring about a large-scale transition to
fully open access.

A key finding of the analysis is that transformative agreements increase
market concentration. Specifically, the three largest commercial
publishers Elsevier, Springer Nature, and Wiley dominate, particularly
in regard to open access provided through transformative agreements.
Together, the three publishers accounted for three-fourths of open
access articles through transformative agreements, despite publishing
less than half of the total publication volume published between 2018
and 2022. This observation aligns with previous research on shifts in
the publishing market following the introduction of funding
opportunities for hybrid open access (Butler et al., 2023; Jahn \&
Tullney, 2016; Shu \& Larivière, 2023). Additionally, the results
confirm previously observed variations by publisher, with Elsevier
exhibiting a different development than Springer Nature and Wiley
(Butler et al., 2023; Jahn et al., 2022). However, it must be noted that
the focus of transformative agreements on publishers with large hybrid
journal portfolios is intentional (Campbell et al., 2022). Because of
transformative agreements with a few large publishers, national
consortia were able to substantially increase their country's annual
open access article volume (Bosman, Jonge, et al., 2021; Huang et al.,
2020; Pinhasi et al., 2021; Taubert et al., 2023; Wenaas, 2022).

Moreover, this study presents varying levels of open access uptake
through transformative agreements across journals, which can be
attributed to the alignment of authors' affiliations and the
availability of such agreements at their institutions. In line with
previous research findings (Butler et al., 2023; Jahn et al., 2022),
high uptake rates were observed across hybrid journals in the Social
Sciences and Humanities. However, it is important to emphasise that
these hybrid journals do not encompass the entire field. For example, a
study by Khanna et al. (2022) found that approximately 60\% of journals
utilizing the open-source publishing platform Open Journal Systems (OJS)
fall within the Social Sciences and Humanities. In these fields, the
majority of full open access journals are so-called ``Diamond OA
journals'' that do not charge publication fees (Bosman, Frantsvåg, et
al., 2021). Rather, this result can be more accurately attributed to the
substantial proportion of authors from high-income countries who publish
in these hybrid journals included in transformative agreements,
particularly when journals are of local relevance, such as through the
belonging to national societies or regional research scope.

Surprisingly, the total publication output of high-income countries
belonging to the OECD in hybrid journals declined substantially after
peaking in 2020, whereas that of BRICS countries doubled from 2018 to
2022. Because the BRICS expansion can be exclusively attributed to
closed access articles, this development has the potential to hinder the
transition of academic journal publishing to full open access through
transformative agreements and demands discussion. In China, the country
with the highest volume of articles in hybrid journals, limited research
funding to pay for open access is available, with expenditures for open
access publishing already surpassing subscription costs (Shu \&
Larivière, 2023). Furthermore, the focus of Chinese authors on
established journals may contribute to this trend (Zhang et al., 2022).
Although China supports Plan S (Schiermeier, 2018), this is not
reflected in the data in terms of open access uptake in hybrid journals.
As highlighted by Koley \& Lala (2023), India faces similar challenges
in terms of resources to pay for open access. At the same time, access
to research literature is a pressing, which is adressed by the ``Indian
one nation, one subscription'' policy proposal. However, this policy
focuses on centrally negotiated subscriptions and does not entail open
access.

But open access uptake also differs among OECD countries. In the United
States, for example, hybrid open access including transformative
agreements plays a lesser role compared to some European countries. The
smallest share of federally funded research was hybrid open access
between 2017 and 2021 (Schares, 2023). While some university consortia
like the California Digital Library signed transformative agreements,
others attempt to depart from big deals and unbundle large journal
portfolios to address cost increases (Brainard, 2021; Schares, 2022).
Despite the relatively low penetration of transformative agreements in
China and the United States compared to Europe, 22\% of open access in
hybrid journals in 2022 originates from these two countries, indicating
the availability of funding sources for publication fees.

This large-scale study provides first empricial evidence about the
influence of transformative agreements on the transition of hybrid
journals to full open access. However, several limitations needs to be
acknowledged. From a data perspective, estimations of open access
through transformative agreements were established by linking first
author affiliations with publicly available agreement data from
cOAlition S, and not through invoicing data because these data a usually
not shared by publishers. Moreover, this study is unable to account for
the various types of transformative agreements, particularly with
regards to article types and capping. Similarly, assessing the quality
of the OpenAlex and Crossref data used, particularly in terms of
affiliations and article types, in comparison with established
bibliometric databases was beyond the scope of this analysis. Overall,
the methodology is designed in such a way that the results underestimate
rather than overestimate the adoption of open access through
transformative agreements. In addition, the data used in this study are
openly available, along with the code used for this analysis.
